---
title: "New Method of NBA Player Classification"
author: "[Daniel Wolf](daniel.wolf33@gmail.com)"
date: "March 1st, 2016"
output: html_document
---

## Introduction

My goal in this project is to contribute to 
advancing the scope and value of basketball analytics.

The methodology below is adapted from _Basketball Analytics: Spatial
Tracking_ by Stephen Shea. 
As Stephen points out, conventional player positions (i.e. point guard, 
center) are not well-suited for player comparison. Players listed at the 
same position can have very different types. For example, consider Tony Allen vs
Kyle Korver for shooting guards, or DeAndre Jordan vs Andrea Bargnani for
centers.

For this reason, there is a need for a new player classification structure that
can compare players more precisely and effectively by grouping them
based on role and/or skill set.


```{r setup, include=FALSE}

#knitr::opts_chunk$set(fig.width=30, fig.height=20, fig.path='Figs/',
#                      echo=FALSE, warning=FALSE, message=FALSE)
knitr::opts_chunk$set(fig.width=15, fig.height=10, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
library(ggplot2)

```


```{r load_data}

hoops <- read.csv('hoops.csv')
hoops$Player.Name <- hoops$Player
hoops <- data.frame(hoops, row.names="Player")

# Remove players below certain thresholds
mins_threshold = 15
game_threshold = 10
hoops <- subset(hoops, hoops[,"GP"]>=game_threshold & 
                hoops[,"MIN"]>=mins_threshold)


# Evaluate some basic attributes of the dataset
#dim(hoops)
#names(hoops)
#str(hoops)
#row.names(hoops)
#head(hoops)
#summary(hoops)

```



```{r new_field_creation}

# Create new fields that will be used in the "filtration system"

# Passes per Touch
hoops$Passes.per.Touch <- with(hoops, Passes.Made.per.Game / Touches.per.Game)
# Possessions Terminated
hoops$PT <- with(hoops, FGA + AO + SecondAST + FTAst + 0.44*FTA)
# Scoring Turnovers
hoops$STOV <- with(hoops, ((FGA + 0.44*FTA) / PT) * TOV)
# Playmaking Turnovers
hoops$PTOV <- with(hoops, TOV - STOV)
# Playmaking Possessions with TOV split
hoops$PP <- with(hoops, AO + SecondAST + FTAst + PTOV)
# Individual Playmaking Efficiency with TOV split
hoops$IPE <- with(hoops, AST.PTS.Created / PP)
# Close Shots
hoops$Close.Shot.PTS <- with(hoops, Paint.Touch.PTS+Post.Touch.PTS+Elbow.Touch.PTS)
# Percentage of points from Drives
hoops$Dr <- with(hoops, Drive.PTS / PTS + 0.000001)
# Percentage of points from Close Shots
hoops$CS <- with(hoops, Close.Shot.PTS / PTS + 0.000001)
# Percentage of points from Catch and Shoots
hoops$Ca <- with(hoops, Catch.Shoot.PTS / PTS + 0.000001)
# Percentage of points from Pull Ups
hoops$PU <- with(hoops, Pull.Up.PTS / PTS + 0.000001)
# Scoring Versatility Index
hoops$SVI <- with(hoops, -Dr*log2(Dr) - CS*log2(CS) - Ca*log2(Ca) - PU*log2(PU))

# Remove NA values
hoops <- na.omit(hoops)
#dim(hoops)

# Function that will print top and bottom entries based on the chosen field
top.bottom <- function(df, x){
  df <- df[order(df[,x]),]
  print(head(df[,x, drop=FALSE]))
  df <- df[order(-df[,x]),]
  print(head(df[,x, drop=FALSE]))
}

#top.bottom(hoops,"Passes.per.Touch")
#top.bottom(hoops,"PT")
#top.bottom(hoops,"STOV")
#top.bottom(hoops,"PTOV")
#top.bottom(hoops,"PP")
#top.bottom(hoops,"IPE")
#top.bottom(hoops,"Close.Shot.PTS")
#top.bottom(hoops,"Dr")
#top.bottom(hoops,"CS")
#top.bottom(hoops,"Ca")
#top.bottom(hoops,"PU")
#top.bottom(hoops,"SVI")

```


```{r standardize}

# Standardize by converting to per-minute numbers
hoops$Passes.per.Touch <- hoops$Passes.per.Touch / hoops$MIN
hoops$PP <- hoops$PP / hoops$MIN
hoops$FGA3 <- hoops$FGA3 / hoops$MIN
hoops$RAFGA <- hoops$RAFGA / hoops$MIN
hoops$REB.Chances <- hoops$REB.Chances / hoops$MIN
hoops$Opp.FGA.at.Rim <- hoops$Opp.FGA.at.Rim / hoops$MIN
hoops$BLK <- hoops$BLK / hoops$MIN

hoops$PTS <- hoops$PTS / hoops$MIN
hoops$SVI <- hoops$SVI / hoops$MIN
hoops$Opp.FGPer.at.Rim <- hoops$Opp.FGPer.at.Rim / hoops$MIN
hoops$STL <- hoops$STL / hoops$MIN
hoops$IPE <- hoops$IPE / hoops$MIN
hoops$AST <- hoops$AST / hoops$MIN
hoops$FGM3 <- hoops$FGM3 / hoops$MIN

# Function that will standardize using the mean and standard deviation
standardize <- function(df, x){
  average <- mean(df[,x])
  stdev <- sd(df[,x])
  df[,x] <- (df[,x] - average) / stdev
  return(df)
}

hoops <- standardize(hoops, "Passes.per.Touch")
hoops <- standardize(hoops, "PP")
hoops <- standardize(hoops, "FGA3")
hoops <- standardize(hoops, "RAFGA")
hoops <- standardize(hoops, "REB.Chances")
hoops <- standardize(hoops, "Opp.FGA.at.Rim")
hoops <- standardize(hoops, "BLK")
hoops <- standardize(hoops, "PTS")
hoops <- standardize(hoops, "SVI")
hoops <- standardize(hoops, "Opp.FGPer.at.Rim")
hoops <- standardize(hoops, "STL")
hoops <- standardize(hoops, "IPE")
hoops <- standardize(hoops, "AST")
hoops <- standardize(hoops, "FGM3")

#top.bottom(hoops, "Passes.per.Touch")
#top.bottom(hoops, "PP")
#top.bottom(hoops, "FGA3")
#top.bottom(hoops, "RAFGA")
#top.bottom(hoops, "REB.Chances")
#top.bottom(hoops, "Opp.FGA.at.Rim")
#top.bottom(hoops, "BLK")
#top.bottom(hoops, "PTS")
#top.bottom(hoops, "SVI")
#top.bottom(hoops, "Opp.FGPer.at.Rim")
#top.bottom(hoops, "STL")
#top.bottom(hoops, "IPE")
#top.bottom(hoops, "AST")
#top.bottom(hoops, "FGM3")

```


```{r polar_conversion}

# Convert to x,y coordinates for plotting

# x coordinates
hoops$x.Passes.per.Touch = hoops$Passes.per.Touch * cos(4*pi/4)
hoops$x.PP = hoops$PP * cos(4*pi/4)
hoops$x.FGA3 = hoops$FGA3 * cos(2*pi/4)
hoops$x.RAFGA = hoops$RAFGA * cos(6*pi/4)
hoops$x.REB.Chances = hoops$REB.Chances * cos(0*pi/4)
hoops$x.Opp.FGA.at.Rim = hoops$Opp.FGA.at.Rim * cos(0*pi/4)

hoops$x.BLK = hoops$BLK * cos(7*pi/4)
hoops$x.PTS = hoops$PTS * cos(1*pi/4)
hoops$x.SVI = hoops$SVI * cos(1*pi/4)
hoops$x.Opp.FGPer.at.Rim = hoops$Opp.FGPer.at.Rim * cos(7*pi/4)
hoops$x.STL = hoops$STL * cos(5*pi/4)
hoops$x.IPE = hoops$IPE * cos(3*pi/4)
hoops$x.AST = hoops$AST * cos(3*pi/4)
hoops$x.FGM3 = hoops$FGM3 * cos(2*pi/4)

# y coordinates
hoops$y.Passes.per.Touch = hoops$Passes.per.Touch * sin(4*pi/4)
hoops$y.PP = hoops$PP * sin(4*pi/4)
hoops$y.FGA3 = hoops$FGA3 * sin(2*pi/4)
hoops$y.RAFGA = hoops$RAFGA * sin(6*pi/4)
hoops$y.REB.Chances = hoops$REB.Chances * sin(0*pi/4)
hoops$y.Opp.FGA.at.Rim = hoops$Opp.FGA.at.Rim * sin(0*pi/4)

hoops$y.BLK = hoops$BLK * sin(0*pi/4)
hoops$y.PTS = hoops$PTS * sin(1*pi/4)
hoops$y.SVI = hoops$SVI * sin(1*pi/4)
hoops$y.Opp.FGPer.at.Rim = hoops$Opp.FGPer.at.Rim * sin(7*pi/4)
hoops$y.STL = hoops$STL * sin(5*pi/4)
hoops$y.IPE = hoops$IPE * sin(3*pi/4)
hoops$y.AST = hoops$AST * sin(3*pi/4)
hoops$y.FGM3 = hoops$FGM3 * sin(2*pi/4)

hoops$x.coord <- with(hoops,(x.Passes.per.Touch + x.PP + x.FGA3 +
                       x.RAFGA + x.REB.Chances +
                       x.Opp.FGA.at.Rim + x.BLK +
                       x.PTS + x.SVI + x.Opp.FGPer.at.Rim +
                       x.STL + x.IPE + x.AST + x.FGM3) / 14 )
hoops$y.coord <- with(hoops,(y.Passes.per.Touch + y.PP + y.FGA3 +
                       y.RAFGA + y.REB.Chances +
                       y.Opp.FGA.at.Rim + y.BLK +
                       y.PTS + y.SVI + y.Opp.FGPer.at.Rim +
                       y.STL + y.IPE + y.AST + y.FGM3) / 14 )


```

## Visual Classification

Consider the chart below. The location of each player is determined by 
his play, as measured by a select set of metrics (shown in 
__<span style="color:blue">blue</span>__). 
Each player is "pulled" toward the metrics that best define that
player. For example, versatile scorers are pulled to the top-right 
(see __Kevin Durant__, __Dirk Nowitzki__), 3-point specialists are pulled up 
(see __JJ Redick__, __Steph Curry__), playmakers are pulled to the left 
(see __Rajon Rondo__, __Chris Paul__), and defensive big men are pulled to the 
bottom-right (see __DeAndre Jordan__, __Andre Drummond__).

> __The key insight is that the closer a player is to any other player on 
the chart, the more similar they are based on role and skill set.__


```{r charts}

ggplot(aes(x = x.coord, y = y.coord), data = hoops) +
  geom_point(aes(x=x.coord, y = y.coord), alpha = 0) +
  theme(axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank()) +
  geom_text(aes(label=hoops$Player), size = 4) + 
  annotate("text", x=.45, y=.45, label= "Points,\nScoring Versatility", 
           size = 8, color = "blue")+
  annotate("text", x=0, y=0.6, label= "3FGA,\n3FGM", 
           size = 8, color = "blue")+
  annotate("text", x=-.45, y=0.45, label= "Individual Playmaking Efficiency,\nAssists", 
           size = 8, color = "blue")+
  annotate("text", x=0.59, y=0.09, label= "Rebound\nChances,\nOpp. FGA at Rim", 
           size = 8, color = "blue")+
  annotate("text", x=-.6, y=0.09, label= "Passes per Touch,\nPlaymaking Poss.",
           size = 8, color = "blue")+
  annotate("text", x=-.4, y=-.4, label= "Steals", 
           size = 8, color = "blue")+
  annotate("text", x=0, y=-.6, label= "Restricted Area FGA", 
           size = 8, color = "blue")+
  annotate("text", x=0.59, y=-.5, label= "Blocks,\nOpp. FG% at Rim", 
           size = 8, color = "blue")


```

## Value Add

In this way, very precise player classifications can be 
defined given that players next to each other are similar.
Here are three potential ways this approach could add value:

1. Support the "eye test" with data. Scouts and front offices consistently 
evaluate a player by referencing "similar" players to get a better 
understanding of how a player might fit in a system or a role.
2. Once a team need is identified, use this method to find players that 
fit that type and skill set. Also, if
a target player has been identified, quickly identify other alternatives.
3. Taking this analysis one step further could be used to determine an 
optimal combination of player types to create a winning team.



## Metrics

Some of these metrics characterize a player's role, while 
others describe a player's skill within that role. 
The set of metrics used in this exercise can easily be adjusted depending 
on the intended use. All metrics 
were converted to __per-minute__ and __normalized__ in order to more
effectively compare player ability while they are on the floor. Players 
with less than 10 GP and 15 MPG in the 2015-2016 season have been removed.
There are a total of __292 players__ included in this analysis.

__Opportunity Metrics: Identify the player's ROLE (e.g. big or small)__

* 3-point Field Goals Attempted
* Restricted Area Field Goals Attempted
* Rebounding Chances
* Opponent Field Goal Attempts at Rim
* Passes per Touch
* Playmaking Possessions (how many opportunities does a player have to put 
other players in a position to score)

__Performance Metrics: Identify the player's SKILL SET__

* Points
* Steals
* Assists
* Blocks
* 3-point Field Goals Made
* Opponent Field Goal Percentage at Rim
* Individual Playmaking Efficiency (how efficiently can the player
put other players in a position to score)
* Scoring Versatility Index (how evenly distributed is scoring between pull-ups,
close shots, drives, and catch and shoot)

All of the metrics needed for this report either come directly
from NBA.com or are calculated using statistics from NBA.com.


```{r closest_players}

# This function will find the closest players for any player based on the metrics
closest.players <- function(df, player){
  
  # Pull out the player names and x,y coordinates to a new data frame
  xy = data.frame(Player.Name = df$Player.Name, 
                  x.coord = df$x.coord, 
                  y.coord = df$y.coord)
  
  # Replicate the rows in order to prep for calculations
  xy.rep.1 <- xy[rep(row.names(xy), nrow(xy)),]
  xy.rep.2 <- xy.rep.1[order(xy.rep.1[,"Player.Name"]),]
  
  # Rename the columns that will be compared to
  names(xy.rep.1)[names(xy.rep.1) == 'Player.Name'] <- 'Player.comp'
  names(xy.rep.1)[names(xy.rep.1) == 'x.coord'] <- 'x.comp'
  names(xy.rep.1)[names(xy.rep.1) == 'y.coord'] <- 'y.comp'
  
  # Combine the two dataframes together for comparison
  closest <- cbind(xy.rep.2, xy.rep.1)
  
  # Calculate the distance between the two points
  closest$distance <- with(closest, sqrt((x.coord-x.comp)^2 + (y.coord-y.comp)^2))
  
  # Sort low to high in terms of Player and Distance
  closest <- closest[order(closest[,"Player.Name"],closest[,"distance"]),]
  
  # Subset to only output the player of interest
  closest = subset(closest, Player.Name == player)
  
  # Print the results
  print(head(closest[,"Player.comp",drop=FALSE],8))
  
}


# Enter a player name to find the closest players
#closest.players(hoops, "Dennis Schroder")
#closest.players(hoops, "Thabo Sefolosha")
#closest.players(hoops, "Kent Bazemore")
#closest.players(hoops, "Jeff Teague")
#closest.players(hoops, "Kyle Korver")
#closest.players(hoops, "Paul Millsap")
#closest.players(hoops, "Al Horford")
#closest.players(hoops, "Mike Scott")
#closest.players(hoops, "Kirk Hinrich")
#closest.players(hoops, "Kris Humphries")
#closest.players(hoops, "Tim Hardaway Jr.")
#closest.players(hoops, "Tiago Splitter")

```


```{r output}

#dim(hoops)
write.csv(hoops, "hoops_out.csv")

```

## Focus On: Hawks


Players closest to Jeff Teague:

* Ty Lawson
* Tyreke Evans
* Mario Chalmers

Players closest to Kent Bazemore:

* Dante Cunningham
* James Johnson
* Matt Barnes

Players closest to Kyle Korver:

* PJ Hairston
* Anthony Tolliver
* Wesley Matthews

Players closest to Paul Millsap:

* Jerami Grant
* Terrence Jones
* Jared Sullinger

Players closest to Al Horford:

* Shabazz Muhammad
* Jon Leuer
* Ersan Ilyasova

Players closest to Dennis Schroder:

* Toney Douglas
* Jameer Nelson
* Steve Blake

Players closest to Thabo Sefolosha:

* PJ Tucker
* Tayshaun Prince
* Andre Roberson

Players closest to Tim Hardaway Jr.:

* Isaiah Thomas
* Manu Ginobili
* Jerryd Bayless

Players closest to Tiago Splitter:

* Tony Allen (interesting)
* Trevor Booker
* Rondae Hollis-Jefferson 
* Larry Nance Jr.

Players closest to Kris Humphries:

* Andrew Nicholson
* Ersan Ilyasova
* Kristaps Porzingis


For more information or to discuss these topics in more depth, 
feel free to reach out to Daniel Wolf at daniel.wolf33@gmail.com. This 
project was completed using R programming.
