---
title: "Measuring Football Defensive Stretch"
author: "[Daniel Wolf](daniel.wolf33@gmail.com)"
date: "March 10, 2016"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(fig.width=15, fig.height=10, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
library(ggplot2)
library(dplyr)

```


```{r load_data}

football.orig <- read.csv('UnderArmor2016.csv')

# Evaluate some basic attributes of the dataset
#dim(football)
#names(football)
#str(football)
#row.names(football)
#head(football)
#summary(football)

football.orig$defense <- ifelse(football.orig$player < 100,0,1)

# New unique identifier fields
football.orig <- within(football.orig, Play.Frame.D <- paste(play, frame, defense, sep='_'))
football.orig <- within(football.orig, Play.Frame <- paste(play, frame, sep='_'))
football.orig <- within(football.orig, Play.D <- paste(play, defense, sep='_'))

football <- football.orig

```


```{r centroid}

# Create new columns to identify the centroid for each Play/Frame/Defense
football.xz.mean.per.frame <- summarize(group_by(football, Play.Frame.D), 
                                    x.mean.per.frame = mean(x), 
                                    z.mean.per.frame = mean(z))
football <- merge(football,football.xz.mean.per.frame)

# Create new columns to identify the average centroid for each Play/Defense
football.xz.mean.per.play <- summarize(group_by(football, Play.D), 
                                   x.mean.per.play = mean(x), 
                                   z.mean.per.play = mean(z))
football <- merge(football,football.xz.mean.per.play)


```

```{r ball_position}

# Create new columns to identify the ball position for each Play/Frame
football.HasBall <- subset(football, hasball == 1)
football.HasBall$x.ball.per.frame <- football.HasBall$x
football.HasBall$z.ball.per.frame <- football.HasBall$z
football.HasBall <- football.HasBall[,c("Play.Frame", 
                                        "x.ball.per.frame", 
                                        "z.ball.per.frame")]

# If there are NAs with 2nd method, then there are frames that do not have a HasBall ticker
# So you do not know the position of the ball
# Merge defaults to inner join
football <- merge(football,football.HasBall)
#football <- merge(x = football, y = football.HasBall, by = "Play.Frame", all.x = TRUE)

# Create new columns to identify the average ball position for each Play/Defense
football.xz.ball.per.play <- summarize(group_by(football, Play.D), 
                                   x.ball.per.play = mean(x.ball.per.frame), 
                                   z.ball.per.play = mean(z.ball.per.frame))
football <- merge(football,football.xz.ball.per.play)



```


```{r dda_ddb}

# Calculate each player's distance from average
football$dda.using.mean.per.frame <- with(football, 
                           sqrt((x-x.mean.per.frame)^2 + (z-z.mean.per.frame)^2))
football$dda.using.mean.per.play <- with(football, 
                          sqrt((x-x.mean.per.play)^2 + (z-z.mean.per.play)^2))

# Calculate each player's distance from the ball
football$ddb.using.ball.per.frame <- with(football, 
                     sqrt((x-x.ball.per.frame)^2 + (z-z.ball.per.frame)^2))
football$ddb.using.ball.per.play <- with(football, 
                     sqrt((x-x.ball.per.play)^2 + (z-z.ball.per.play)^2))

#----------------------------------------------------------------------------------
# Group and summarize by Play/Frame/Defense to get dda average per frame
football.dda.mean.per.frame <- summarize(group_by(football, Play.Frame.D), 
                                dda.mean.per.frame = mean(dda.using.mean.per.frame))
football <- merge(football,football.dda.mean.per.frame)

# Group and summarize by Play/Frame/Defense to get ddb average per frame
football.ddb.mean.per.frame <- summarize(group_by(football, Play.Frame.D), 
                                ddb.mean.per.frame = mean(ddb.using.ball.per.frame))
football <- merge(football,football.ddb.mean.per.frame)

#----------------------------------------------------------------------------------
# Group and summarize by Play/Defense to get dda per play (empiric since using players)
football.dda.mean.per.play.empiric <- summarize(group_by(football, Play.D), 
                               dda.mean.per.play.empiric = mean(dda.using.mean.per.play))
football <- merge(football,football.dda.mean.per.play.empiric)

# Group and summarize by Play/Defense to get ddb per play (empiric since using players)
football.ddb.mean.per.play.empiric <- summarize(group_by(football, Play.D), 
                               ddb.mean.per.play.empiric = mean(ddb.using.ball.per.play))
football <- merge(football,football.ddb.mean.per.play.empiric)

#----------------------------------------------------------------------------------

# Group and summarize by Play/Defense to get dda per play
# This equals the average of dda.mean.per.frame, so it is an average of the frame averages
football.dda.mean.per.play.rollup <- summarize(group_by(football, Play.D), 
                               dda.mean.per.play.rollup = mean(dda.using.mean.per.frame))
football <- merge(football,football.dda.mean.per.play.rollup)

# Group and summarize by Play/Defense to get ddb per play
# This equals the average of ddb.mean.per.frame, so it is an average of the frame averages
football.ddb.mean.per.play.rollup <- summarize(group_by(football, Play.D), 
                               ddb.mean.per.play.rollup = mean(ddb.using.ball.per.frame))
football <- merge(football,football.ddb.mean.per.play.rollup)


```


```{r firstlast_frame_1}

#head(football)

# Create new columns to identify which frames are the first and last frames for each play
# Switch to ORIGINAL football if causing issues
football.frames <- summarize(group_by(football, Play.D),
                             min.frame = min(frame),max.frame = max(frame))
football <- merge(football,football.frames)
#write.csv(football.frames, "football_frames.csv")


football$first.frame <- ifelse(football$min.frame == football$frame,1,0)
football$last.frame <- ifelse(football$max.frame == football$frame,1,0)
football$min.frame <- NULL
football$max.frame <- NULL
football.first.defense.frame <- subset(football, first.frame == 1 & defense == 1)
football.last.defense.frame <- subset(football, last.frame == 1 & defense == 1)

```

```{r firstlast_frame_2}
football.first.defense.frame <- football.first.defense.frame[,c("play", 
                                        "dda.mean.per.frame", 
                                        "ddb.mean.per.frame")]


# Rename columns to first and last
names(football.first.defense.frame)[names(football.first.defense.frame) == 
                                      'dda.mean.per.frame'] <- 'dda.mean.first.frame'
names(football.first.defense.frame)[names(football.first.defense.frame) == 
                                      'ddb.mean.per.frame'] <- 'ddb.mean.first.frame'


football.last.defense.frame <- football.last.defense.frame[,c("play", 
                                        "dda.mean.per.frame", 
                                        "ddb.mean.per.frame")]
# Rename columns to first and last
names(football.last.defense.frame)[names(football.last.defense.frame) == 
                                      'dda.mean.per.frame'] <- 'dda.mean.last.frame'
names(football.last.defense.frame)[names(football.last.defense.frame) == 
                                      'ddb.mean.per.frame'] <- 'ddb.mean.last.frame'

football.firstlast.frame <- merge(football.first.defense.frame,football.last.defense.frame)

# Remove duplicates (should end up with 63 rows for 63 plays)
football.firstlast.frame <-
  football.firstlast.frame[!duplicated(football.firstlast.frame), ]



```


```{r streamlined_to_play_defense}

# Streamline to defense only
football.defense <- subset(football, defense == 1)

# Streamline to play-level fields (remember this is defense only)
football.by.play <- football.defense[,c("play","team","playtype","fdl",
                                "los","down",
                                "dda.mean.per.play.empiric",
                                "ddb.mean.per.play.empiric",
                                "dda.mean.per.play.rollup",
                                "ddb.mean.per.play.rollup")]

# Remove duplicates (should end up with 63 rows for 63 plays)
football.by.play <- football.by.play[!duplicated(football.by.play), ]

# Bring in dda and ddb for first and last frame
football.by.play <- merge(football.by.play,football.firstlast.frame)

# Calculate increase in dda and ddb
football.by.play$dda.increase <- football.by.play$dda.mean.last.frame -
                                  football.by.play$dda.mean.first.frame
football.by.play$ddb.increase <- football.by.play$ddb.mean.last.frame -
                                  football.by.play$ddb.mean.first.frame



```


```{r prep_for_plotting}

football.by.playtype <- summarize(group_by(football.by.play, playtype), 
                                    mean.start.dda = mean(dda.mean.first.frame), 
                                    mean.start.ddb = mean(ddb.mean.first.frame),
                                    mean.finish.dda = mean(dda.mean.last.frame),
                                    mean.finish.ddb = mean(ddb.mean.last.frame)
                                  )
#football.by.playtype
dda.start <- football.by.playtype[,c("playtype","mean.start.dda")]
ddb.start <- football.by.playtype[,c("playtype","mean.start.ddb")]
dda.finish <- football.by.playtype[,c("playtype","mean.finish.dda")]
ddb.finish <- football.by.playtype[,c("playtype","mean.finish.ddb")]

dda.start$phase <- rep("Start of Play",nrow(dda.start))
ddb.start$phase <- rep("Start of Play",nrow(ddb.start))
dda.finish$phase <- rep("End of Play",nrow(dda.finish))
ddb.finish$phase <- rep("End of Play",nrow(ddb.finish))

dda.start$metric <- rep("dda",nrow(dda.start))
ddb.start$metric <- rep("ddb",nrow(ddb.start))
dda.finish$metric <- rep("dda",nrow(dda.finish))
ddb.finish$metric <- rep("ddb",nrow(ddb.finish))

names(dda.start)[names(dda.start) == 'mean.start.dda'] <- 'distance'
names(ddb.start)[names(ddb.start) == 'mean.start.ddb'] <- 'distance'
names(dda.finish)[names(dda.finish) == 'mean.finish.dda'] <- 'distance'
names(ddb.finish)[names(ddb.finish) == 'mean.finish.ddb'] <- 'distance'

dda.start <- within(dda.start, matching <- paste(playtype, metric, sep='_'))
ddb.start <- within(ddb.start, matching <- paste(playtype, metric, sep='_'))
dda.finish <- within(dda.finish, matching <- paste(playtype, metric, sep='_'))
ddb.finish <- within(ddb.finish, matching <- paste(playtype, metric, sep='_'))

football.ready <- rbind(dda.start,ddb.start)

football.ready <- rbind(football.ready, dda.finish)
football.ready <- rbind(football.ready, ddb.finish)

x <- c(0,1,2,3,4,5,6,7)
y <- c("Pass","Rush/Run","Kick/Punt/FG",
                                "Incomplete","Sack","Penalty",
                                "Interception","Touchdown")
play.type.mapping <- data.frame(x,y)
names(play.type.mapping) = c("playtype","Play.Type")


football.ready <- merge(football.ready,play.type.mapping)


football.ready$playtype <- ifelse(football.ready$metric=="dda",
                                    football.ready$playtype + 0.15,
                                    football.ready$playtype - 0.15)


```



```{r chart}

ggplot(aes(x=playtype,y=distance,group=matching), data = football.ready) +
  geom_point(aes(color = phase),
             pch="|", size = 12) +
  geom_line(aes(linetype=metric), 
            size = 0.75) +
  scale_color_manual(values=c("red", "green4")) +
  scale_x_continuous(breaks=c(0,1,2,3,4,5,6,7), 
                   labels=c("Pass","Rush/Run","Kick/Punt/FG",
                                "Incomplete","Sack","Penalty",
                                "Interception","Touchdown")) +
  scale_y_continuous(breaks=c(20,30,40,50,60,70)) +
  ggtitle("Changes in dda and ddb for Different Play Types") +
  xlab("Play Type") + 
  ylab("Average defensive distance in feet") +
  theme(axis.text=element_text(size=20),
        axis.title=element_text(size=22,face="bold"),
        legend.title=element_blank(),
        legend.text=element_text(size=18),
        plot.title = element_text(size = 25),
        legend.key.size = unit(1.5,"cm")
        ) +
  coord_flip()

```



### Definition

* __Defense Distance from Average (dda)__ represents the average of each defender's 
distance from the centroid (i.e. average location) of the defense.
* __Defense Distance from Ball (ddb)__ represents the average of each defender's distance
from the ball.

#### About

I created this during the ESPN Hackathon at the 2016 Sloan Sports Analytics Conference. 
The goal was measure the stretch of a football defense. I applied two metrics, dda and ddb,
to measure the defensive stretch in two different ways.
The data source is player tracking data from the Under Armor high school all-star game.
For R code used to calculate and visualize, you can visit http://github.com/dwolfATL/Sports-Analytics.
