---
title: "A View of Big Man 'Care Factor'"
author: "[Daniel Wolf](daniel.wolf33@gmail.com)"
date: "March 2nd, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=9, fig.height=6,
                      echo=FALSE, warning=FALSE, message=FALSE)

library(ggplot2)
library(RColorBrewer)

```



```{r visualize}

df <- read.csv('playersummary.csv')

# Sort by the key metric for plotting
df <- transform(df,name_end = reorder(name_end, -mean_bask_diff))

colourCount = length(unique(df$name_end))

ggplot(df, aes(x=name_end,y=mean_bask_diff,color='black',fill=name_end)) +
  geom_bar(stat='identity') +
  scale_fill_manual(values = colorRampPalette(brewer.pal(9, "Greens"))(colourCount)) +
  xlab('Player Name') +
  ylab('Change in distance to basket during shot attempt (feet)') +
  ggtitle('How hard are the bigs crashing the boards?') +
  theme(legend.position='none') + 
  theme(plot.title = element_text(size=22)) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))


```

#### Approach

* Measured each big man's **change in their distance to the basket** 
while the ball is in the air during a shot attempt
* All scenarios where they started between 10 to 20 feet from the basket. 
There are 1,231 samples across 23 big men for these 5 games.
* Regardless of whether they actually get the rebound. This metric 
is based on effort as opposed to skill.

#### Key Takeaways

* On average, Tristan Thompson gets 4 feet closer to the basket while the ball is in the air!
* DeMarcus Cousins and Roy Hibbert get further from the basket in the same scenarios

#### About

I created this plot during the ESPN Hackathon at the 2017 Sloan Sports Analytics Conference. 
The Hackathon prompt was to "Measure the Immeasurable", so I attempted to quantify
how much effort the big men are putting into their rebounding.
The data source is SportVU player tracking from 5 full NBA games in the 2015-16 season. 
For R code used to calculate and visualize, you can visit 
[http://github.com/dwolfATL/Sports-Analytics](http://github.com/dwolfATL/Sports-Analytics).

